---
title: "How Inefficient is my portofolio?"
author: "Jeffrey Waring"
date: "10 OCT 2016"
output: html_document
---

```{r, echo=FALSE}
knitr::opts_chunk$set(error = TRUE)
```
  
**Abstract**  
Markowitz detailed, in his work on portfolio theory, that optimal combinations of assets exist.  That is, given a level of risk, there is a combination of investments that will have the greatest return.  These possible combinations form a "frontier", or "Markowitz Bullet".  Ideally, ones portfolio should lie on this frontier (that is, you're getting the highest return for your desired level of risk).  An investor can then leverage some assets, and short others, to exceed even the frontier.   

Currently, I have a random assortment of mutual funds in my IRA and 401k.  Likewise, I have a few securities, purchased on a whim or through my employers stock purchase program.  I am curious as to how "inefficient" my current asset distribution is.

To that end, I first created a list of my current assets, and their relative weights in my portfolio:

```{r}
stocks <- c(
  "DODFX" = .063,
  "VIIIX" = .223,
  "VMCPX" = .055,
  "VSCPX" = .027,
  "VBMPX" = .063,
  "JEC"   = .178,
  "GOOG"   = .047,
  "AAPL"  = 0.0077,
  "VMGMX" = 0.119,
  "VWINX" = 0.030)
```

Using the package 'stockPortfolio', I will retrieve the returns and volatilities of each asset
```{r}
library(stockPortfolio)
stockData <- getReturns(names(stocks),start="2013-01-01",end=NULL,freq="month")
```

Next, using the quadprog package, the returns and risk of a variety of portfolios is calculated. 
```{r echo=FALSE}
library(knitr)
library(ggplot2) # Used to graph efficient frontier
library(reshape2) # Used to melt the data
library(quadprog) #Needed for solve.QP

```

And then the plot is created:

```{r echo=FALSE}
library(stockPortfolio) # Base package for retrieving returns
library(ggplot2) # Used to graph efficient frontier
library(reshape2) # Used to melt the data
library(quadprog) #Needed for solve.QP

#### Efficient Frontier function ####
eff.frontier <- function (returns, short="no", max.allocation=NULL,
                          risk.premium.up=.5, risk.increment=.005){
  # return argument should be a m x n matrix with one column per security
  # short argument is whether short-selling is allowed; default is no (short
  # selling prohibited)max.allocation is the maximum % allowed for any one
  # security (reduces concentration) risk.premium.up is the upper limit of the
  # risk premium modeled (see for loop below) and risk.increment is the
  # increment (by) value used in the for loop
  
  covariance <- cov(returns)
  #print(covariance)
  n <- ncol(covariance)
  
  # Create initial Amat and bvec assuming only equality constraint
  # (short-selling is allowed, no allocation constraints)
  Amat <- matrix (1, nrow=n)
  bvec <- 1
  meq <- 1
  
  # Then modify the Amat and bvec if short-selling is prohibited
  if(short=="no"){
    Amat <- cbind(1, diag(n))
    bvec <- c(bvec, rep(0, n))
  }
  
  # And modify Amat and bvec if a max allocation (concentration) is specified
  if(!is.null(max.allocation)){
    if(max.allocation > 1 | max.allocation <0){
      stop("max.allocation must be greater than 0 and less than 1")
    }
    if(max.allocation * n < 1){
      stop("Need to set max.allocation higher; not enough assets to add to 1")
    }
    Amat <- cbind(Amat, -diag(n))
    bvec <- c(bvec, rep(-max.allocation, n))
  }
  
  # Calculate the number of loops
  loops <- risk.premium.up / risk.increment + 1
  loop <- 1
  
  # Initialize a matrix to contain allocation and statistics
  # This is not necessary, but speeds up processing and uses less memory
  eff <- matrix(nrow=loops, ncol=n+3)
  # Now I need to give the matrix column names
  colnames(eff) <- c(colnames(returns), "Std.Dev", "Exp.Return", "sharpe")
  
  # Loop through the quadratic program solver
  for (i in seq(from=0, to=risk.premium.up, by=risk.increment)){
    dvec <- colMeans(returns) * i # This moves the solution along the EF
    sol <- solve.QP(covariance, dvec=dvec, Amat=Amat, bvec=bvec, meq=meq)
    eff[loop,"Std.Dev"] <- sqrt(sum(sol$solution*colSums((covariance*sol$solution))))
    eff[loop,"Exp.Return"] <- as.numeric(sol$solution %*% colMeans(returns))
    eff[loop,"sharpe"] <- eff[loop,"Exp.Return"] / eff[loop,"Std.Dev"]
    eff[loop,1:n] <- sol$solution
    loop <- loop+1
  }
  
  return(as.data.frame(eff))
}

# Find the optimal portfolio
eff.optimal.point <- eff[eff$sharpe==max(eff$sharpe),]

# graph efficient frontier
# Start with color scheme
ealred <- "#7D110C"
ealtan <- "#CDC4B6"
eallighttan <- "#F7F6F0"
ealdark <- "#423C30"

#Current portfolio:

# Create the portfolio using ETFs, incl. hypothetical non-efficient allocation
stocks <- c(
  "DODFX" = .063,
  "VIIIX" = .223,
  "VMCPX" = .055,
  "VSCPX" = .027,
  "VBMPX" = .063,
  "JEC"   = .178,
  "GOOG"   = .047,
  "AAPL"  = 0.0077,
  "VMGMX" = 0.119,
  "VWINX" = 0.030)

stockData <- getReturns(names(stocks),start="2013-01-01",end=NULL,freq="month")

#model stock behavior

model1<-stockModel(stockData)
#get current returns/risk of current allocation:
currentPerformance<-portReturn(model1,stocks)

#now run efficient portfolio:
#Assume no short and 50% alloc. restrictions
eff <- eff.frontier(returns=returns$R, short="no", max.allocation=.5,
                    risk.premium.up=1, risk.increment=.001)
g<-ggplot(eff, aes(x=Std.Dev, y=Exp.Return)) + geom_point(alpha=.1, color=ealdark) +
  geom_point(data=eff.optimal.point, aes(x=Std.Dev, y=Exp.Return, label=sharpe),
             color=ealred, size=5) +
  annotate(geom="text", x=eff.optimal.point$Std.Dev,
           y=eff.optimal.point$Exp.Return,
           label=paste("Risk: ",
                       round(eff.optimal.point$Std.Dev*100, digits=3),"\nReturn: ",
                       round(eff.optimal.point$Exp.Return*100, digits=4),"%\nSharpe: ",
                       round(eff.optimal.point$sharpe*100, digits=2), "%", sep=""),
           hjust=0, vjust=1.2) +
  ggtitle("Efficient Frontier") +
  labs(x="Risk (standard deviation of portfolio)", y="Return") +
  theme(panel.background=element_rect(fill=eallighttan),
        text=element_text(color=ealdark),
        plot.title=element_text(size=24, color=ealred))


t<-geom_point(aes(x=sqrt(currentPerformance$V),y=currentPerformance$R,label=("Current"),color="Red",size=5))
z<-geom_vline(xintercept=sqrt(currentPerformance$V))

#plot fronteir, current allocation, and vertical line through current risk
g+t+z+theme(legend.position="none")+geom_text(aes(label='current'),x=sqrt(currentPerformance$V)+0.003,y=currentPerformance$R+0.001)
```

It is clear that my current portfolio does not lie on the bullet.  I am curious as to which portfolio allocation has the same risk as my actual.  To that end, a vertical line is drawn through the current allocation, and the data.frame of portfolios is subsetted to this level of risk.  The "efficient" portfolio is then:

```{r}

#The portfolio on the frontier that has the same risk profile has the makeup of:
sameRiskEfficient<-eff[which(abs(eff$Std.Dev-sqrt(currentPerformance$V))==min(abs(eff$Std.Dev-sqrt(currentPerformance$V)))),]

sameRiskEfficient
```

It appears that I should increase my holdings of Google, Vanguard's Institutional Index Fund, and Vanguard's Wellesley Income Fund.  For fun, I will track these two portfolios over the next few months, and see which performs better.

The code for this markdown file, as well as the underlying analysis is availabe @ my [githubpage](https://github.com/waringjr/portfolios)
